---
title: 'MATH 1051H S61 - Lecture 13'
output:
   ioslides_presentation:
     font-family: Lato Semibold
     font-import: http://fonts.googleapis.com/css?family=Lato
     widescreen: yes
     css: ../style.css
     fig_caption: yes
---
<style>
citation {
  font-size: 4px;
}
</style>

# Confidence Intervals for t Cases

## Context

**Confidence interval, a general formula**

$$
\text{point estimate} \pm z^\star \cdot SE
$$

Conditions when the point estimate = $\bar{x}$:

* **Independence:** Observations in the sample must be independent
    - random sample/assignment
    -  if sampling without replacement, $n <$ 10% of population
* **Sample size / skew:** $n \ge 30$ and population distribution should not be extremely skewed

**Note:** We will discuss working with samples where $n < 30$ **now**.

## Confidence Intervals, $n < 30$

When we were doing hypothesis tests, and had small samples, do you
remember what we did? We said that the issue was the SE ... and then
we used a t!

That's all we do here. If $n<30$ and you don't know $\sigma$, you use
a $t$ distribution for your confidence interval.

But basically nothing changes ... instead of using $z^\star$, you use
$t^\star$ instead. That's it.

Let's just solve a couple of problems, and see how it works.

## Problem 1: Highway Speeds

Southbound traffic on the I-280 highway near Cupertino, California
had its speed monitored at 3:30pm on a Wednesday. The sample of 12
cars had mean 97.6 km/hr with standard deviation 6.56 km/hr. Test
the highway patrol's claim that the average speed on this highway
at this time of day is lower than the speed limit of 105 km/hr.

Also compute a 99\% confidence interval for the mean speed.

## Problem 1: Information

What do we have from this problem?

* $n=12$
* $\bar{x} = 97.6$
* $s = 6.56$
* $H_A: \mu < 105$
* $\alpha = 0.01$ (confidence: 99\%)

## Problem 1: Statistic

We're doing a hypothesis test:
$$
H_0: \mu = 105 \qquad \text{versus} \qquad H_A: \mu < 105.
$$

Compute the test statistic:

$$
t_\text{test} = \frac{\bar{x} - \mu_0}{SE_{\bar{x}}}
= \frac{97.6 - 105}{\frac{6.56}{\sqrt{12}}} = -3.907676
$$

## Problem 1: ME

When computing a confidence interval, we use the formula
$$
\text{point estimate} \pm t^\star \cdot \text{SE}
$$
where we use $t^\star$ instead of $z^\star$ because $n < 30$. Thus,

$$
\text{ME} = t^\star \cdot \text{SE} = t^\star \cdot 1.893709
$$
```{r}
qt(p = 0.99 + 0.01/2, df = 12 - 1, lower.tail = TRUE)
```

$$
\text{ME} = 3.1058 \cdot 1.8937 = 5.8815.
$$

## Problem 1: Conclusion (Test)

We found a test statistic of -3.9077, so the p-value is (since the alternative
is $\mu < 105$):
```{r}
pt(-3.9077, df = 12 - 1, lower.tail = TRUE)
```
and we reject the null, and conclude that the average speed **is** less than
105 km/hr.

## Problem 1: Conclusion (CI)

Similarly, we can form the 99\% confidence interval, which is
$$
97.6 \pm \text{ME} = 97.6 \pm 5.8815
$$
which is
```{r}
97.6 + c(-1, 1) * 5.8815
```
and thus the 99\% confidence interval for the true mean speed is (91.7, 103.5) km/hr.

## Problem 2: Pennies

Before 1983, US pennies were made with 97\% copper and 3\% zinc.
After 1983, they were converted to 3\% copper and 97\% zinc to make
them cheaper to manufacture. A simple random sample of 25
post-1983 pennies had an average weight of 2.49910g, with standard
deviation 0.01648g. The US Mint specifies that post-1983 pennies should be manufactured
with mean weight 2.500g. At a 95\% level, do you believe that pennies
are actually being manufactured with mean weight of 2.500g?

Compute a 93\% confidence interval for the mean weight of post-1983 pennies.

## Problem 2: Information

What do we have from this problem?

* $n=25$
* $\bar{x} = 2.49910$
* $s = 0.01648$
* $H_A: \mu \neq 2.500$
* $\alpha = 0.07$ (confidence: 93\%)

## Problem 2: Statistic

We're doing a hypothesis test:
$$
H_0: \mu = 2.500 \qquad \text{versus} \qquad H_A: \mu \neq 2.500.
$$

Compute the test statistic:

$$
t_\text{test} = \frac{\bar{x} - \mu_0}{SE_{\bar{x}}}
= \frac{2.49910 - 2.500}{\frac{0.01648}{\sqrt{25}}} = -0.2730583
$$

## Problem 2: ME

When computing a confidence interval, we use the formula
$$
\text{point estimate} \pm t^\star \cdot \text{SE}
$$
where we use $t^\star$ instead of $z^\star$ because $n < 30$. Thus,

$$
\text{ME} = t^\star \cdot \text{SE} = t^\star \cdot 0.003296.
$$
```{r}
qt(p = 0.93 + 0.07/2, df = 25 - 1, lower.tail = TRUE)
```

$$
\text{ME} = 1.896457 \cdot 0.003296 = 0.00625.
$$

## Problem 2: Conclusion (Test)

We found a test statistic of -0.27306, so the p-value is (since the alternative
is $\mu \neq 2.500$):
```{r}
pt(-0.27306, df = 25 - 1, lower.tail = TRUE) * 2
```
and thus we fail to reject the null, and conclude that we do not have evidence
to conclude that the average weight of pennies is anything but 2.5000 g.

## Problem 2: Conclusion (CI)

Similarly, we can form the 93\% confidence interval, which is
$$
2.49910 \pm \text{ME} = 2.49910 \pm 0.00625
$$
which is
```{r}
2.49910 + c(-1, 1) * 0.00625
```
and thus the 93\% confidence interval for the true mean penny weight
is (2.493, 2.505) g.
